\section{Introduction}

\subsection{Population regulation under reduced external threat}

When predator populations are removed or suppressed, prey populations initially increase but sometimes stabilize at lower-than-expected densities. Traditional models (logistic growth, predator-prey dynamics) predict simple density-dependent regulation, but field observations often show complex behavioral and endocrine responses that alter population trajectories in ways not captured by resource-based models alone.

%% TODO: Need citation for predator removal field studies
%% TODO: Need citation for behavioral responses in prey populations

\subsection{Why population trajectories alone are ambiguous}

The null model comparison presented here shows that population trajectories alone are partly non-identifiable: 7 of 11 alternative models produced indistinguishable population outcomes from HSAP on population data. This non-identifiability is a methodological finding with practical consequences --- a field study measuring only population size cannot distinguish endocrine-mediated regulation from simple density dependence.

\subsection{Why endocrine and behavioral observables matter}

For empirical testing, HSAP-like mechanisms require paired endocrine and behavioral observables. Contemporary U.S. endocrine studies and sex-disaggregated arrest statistics provide candidate proxy series for such testing, but they are ecological, confounded, and cannot establish individual-level causation. Their value is to show that the relevant observable classes exist in public datasets and to identify what would need to be measured to test HSAP-like mechanisms empirically.

%% TODO: Need citation for NHANES hormone panels
%% TODO: Need citation for UCR arrest statistics

\subsection{Scope and contributions}

We present HSAP as a falsifiable simulation model with three contributions: (1) a computational framework linking individual endocrine state to population dynamics through behavioral feedbacks; (2) a showing that identical resource conditions produce qualitatively different population outcomes depending on behavioral and endocrine parameters; and (3) an honest null model comparison showing that population trajectories alone are insufficient for mechanism identification, with external proxy data identifying the observable classes required for empirical testing.

HSAP does not model human sexual orientation; non-reproductive sexual or social behavior is treated only as species-specific ethological behavior.
