\section{Methods}
\label{sec:methods}

We follow the ODD (Overview, Design concepts, Details) protocol for describing individual-based models \citep{grimm2006}.

\subsection{Purpose}
\label{sec:purpose}

HSAP tests whether endocrine-linked behavioral feedbacks can generate population stabilization under reduced external threat. The model proposes that low predator pressure triggers a male testosterone downshift, reducing male aggression and reproductive effort, while female aggression increases under reduced male competition. These behavioral changes, combined with density-dependent reproductive restraint, can stabilize populations below carrying capacity through endocrine-mediated pathways rather than resource limitation alone.

\subsection{Agent Architecture}
\label{sec:agent}

Each agent is characterized by sex (assigned 50/50 at birth), age (in model time steps; 10 steps = 1 model-year), and a suite of physiological and behavioral state variables. Agents are initialized with age drawn uniformly from $[5, 40)$ and energy, health, and rank set to 1.0, 1.0, and 0.5 respectively.

\paragraph{Hormonal state.}
Each agent carries three hormonal axes: testosterone, estrogen, and cortisol. Male testosterone is initialized near 1.0 (base value) with Gaussian noise (SD=0.15); female testosterone near 0.3 (SD=0.05). Estrogen is initialized near 1.0 for females and 0.3 for males. Cortisol is initialized near 0.5 for both sexes (SD=0.1). All hormonal values are clamped to $[0, 1]$ at each time step.

\paragraph{Behavioral state.}
Derived from hormones and environment at each step: aggression tendency, maternal defense tendency (females), mating drive, fertility, and offspring survival probability. These are recomputed every time step; agents do not carry fixed behavioral types.

\paragraph{Social state.}
Each agent has a dominance rank (clamped $[0, 1]$), territory quality, cooperation bonus (accumulated from cooperative actions, capped at 0.3), and a pregnancy flag with gestation countdown (5 steps).

\subsection{Environment}
\label{sec:environment}

The environment provides five static parameters and several dynamic state variables.

\paragraph{Static parameters} (set per scenario):
\begin{itemize}
\item \texttt{resource\_abundance} (0 to 1): baseline resource availability
\item \texttt{predator\_pressure} (0 to 1): external threat level
\item \texttt{disease\_pressure} (0 to 1): disease prevalence
\item \texttt{space\_constraint} (0 to 1): controls how quickly territory availability declines with density
\item \texttt{carrying\_capacity} (integer): $K$ for density computation
\end{itemize}

\paragraph{Dynamic state} (updated each step):
\begin{itemize}
\item \texttt{resource\_abundance}: regenerates toward maximum, consumed by population; modulated by seasonality
\item \texttt{territory\_availability}: $\max(0, 1 - \rho \cdot S)$, where $\rho = N / K$
\item \texttt{predation\_risk}: $P \cdot (1 + 0.5 \cdot \rho)$
\item \texttt{disease\_risk}: $D \cdot (1 + 0.5 \cdot \rho)$
\item \texttt{resource\_scarcity}: $\max(0, 1 - R)$, used in cortisol, aggression, and mortality equations
\end{itemize}

\subsection{Endocrine Update}
\label{sec:endocrine}

Hormones are recomputed for every agent at each time step, creating the causal feedback loop that defines HSAP.

\paragraph{Male testosterone} is driven by five factors:

\begin{equation}
T_{m,t+1} = \min\left(T_{\max},\; T_{\text{base}} + 0.3 \cdot \text{rank} + 0.4 \cdot M_d - 0.3 \cdot \rho - 0.4 \cdot (1 - P) - 0.2 \cdot (1 - \text{health})\right)
\label{eq:male_T}
\end{equation}

The term $-0.4 \cdot (1 - P)$ is the critical endocrine feedback: when predator pressure is high, the term is near zero (testosterone is not reduced); when predator pressure is low, the term subtracts up to 0.4 (testosterone is downshifted). This is the mechanism through which threat reduction propagates to behavioral change.

\paragraph{Female testosterone} responds to resource scarcity and density:

\begin{equation}
T_{f,t+1} = \min\left(T_{\max},\; 0.3 + 0.15 \cdot \text{rank} + 0.4 \cdot (1 - R) + 0.3 \cdot \rho\right)
\label{eq:female_T}
\end{equation}

\paragraph{Cortisol} (both sexes) integrates density, rank, health, predation, and resource scarcity:

\begin{equation}
C = 0.5 + 0.1 \cdot \rho - 0.3 \cdot \text{rank} + 0.5 \cdot (1 - \text{health}) + 0.2 \cdot P + 0.3 \cdot (1 - R)
\label{eq:cortisol}
\end{equation}

\paragraph{Estrogen} (females) is suppressed by cortisol:

\begin{equation}
E = 1.0 - 0.3 \cdot C
\label{eq:estrogen}
\end{equation}

\subsection{Behavioral Output}
\label{sec:behavior}

From hormones, agents derive:

\begin{itemize}
\item \textbf{Aggression}: base $= T \cdot 0.5 + C \cdot 0.2$, with sex-specific additions (females: offspring defense + resource competition; males: dominance bonus)
\item \textbf{Fertility}: base $= 0.5$, modified by hormones, age, density-dependent reproductive restraint (activates above density 0.7), and territory availability
\item \textbf{Mating drive}: males $= T \cdot 0.5 + 0.3$; females $= E \cdot 0.4 + 0.2$
\item \textbf{Offspring survival probability}: affected by testosterone, cortisol, density, and resource scarcity
\end{itemize}

\paragraph{Behavioral decision process.}
At each time step, agents select one action from a fixed repertoire: forage, dominance\_event, disperse, withdraw, cooperate, guard\_offspring. Selection follows a priority-based stochastic process:

\begin{enumerate}
\item \textbf{Forage} (deterministic): if energy $< 0.3$
\item \textbf{Guard offspring} (deterministic): if female with dependent offspring and cortisol $> 0.6$
\item \textbf{Dominance event}: probability $= \text{aggression\_tendency} \times 0.3$
\item \textbf{Disperse}: if density $> 0.6$, age $> 15$, probability $= 0.1$
\item \textbf{Withdraw}: if cortisol $> 0.7$, probability $= 0.3 + \eta \cdot \beta_{\text{withdraw}}$
\item \textbf{Cooperate}: if energy $> 0.7$, probability $= 0.1$
\item \textbf{Forage} (default)
\end{enumerate}

\begin{table}[h]
\centering
\caption{Action effects on agent state}
\label{tab:actions}
\begin{tabular}{lll}
\toprule
Action & Energy & Rank \\
\midrule
Forage (success) & $+0.2$ to $+0.4$ & --- \\
Forage (fail) & $-0.05(1+\rho)$ & --- \\
Dominance (win) & $-0.15$ & $+0.05$ \\
Dominance (lose) & $-0.225$ & $-0.05$ \\
Disperse & $-0.3$ & $-0.1$ \\
Withdraw & $+0.05$ & $-0.02$ \\
Cooperate & $+0.05$ & --- \\
Guard offspring & $-0.1$ & --- \\
\bottomrule
\end{tabular}
\end{table}

All action effects are clamped to $[0, 1]$.

\subsection{Reproduction}
\label{sec:reproduction}

Mating occurs when a randomly selected male (probability proportional to mating drive) encounters a fertile female above a fertility gate (default 0.3). Both must have energy above a threshold (0.4) and be in reproductive age (10--60 model-years). On successful mating, the female becomes pregnant with gestation time of 5 steps.

At birth, litter size is drawn from a normal distribution (mean=4.0, SD=1.0, floored to 1). Each pup survives with probability determined by the mother's offspring survival probability. The mother pays an energy cost of 0.3. Newborns face an additional mortality check.

\paragraph{Infanticide} triggers when density exceeds 0.8, with probability proportional to agent aggression and density above threshold. \paragraph{Offspring neglect} triggers when density exceeds 0.7 or cortisol exceeds 0.7, removing one offspring.

\subsection{Mortality}
\label{sec:mortality}

Death probability per time step is computed as:

\begin{equation}
P_{\text{death}} = \min\left(1,\; (P_{\text{base}} + P_{\text{pred}} + P_{\text{disease}} + P_{\text{starvation}} + P_{\text{injury}} + P_{\text{old}} + P_{\text{crowd}}) \times \tau \times M\right)
\label{eq:mortality}
\end{equation}

where $\tau = 0.1$ (age increment per step) and $M$ is the mortality multiplier (1.0 normally, 0.4 during post-sink recovery).

\begin{table}[h]
\centering
\caption{Mortality components}
\label{tab:mortality}
\begin{tabular}{ll}
\toprule
Component & Formula \\
\midrule
Base & 0.02 (constant) \\
Predation & $0.05 \cdot P_{\text{risk}} \cdot (1 - 0.02 \cdot (1-R)) \cdot \text{modifiers}$ \\
Disease & $0.03 \cdot D_{\text{risk}} \cdot (1 + 0.5 \cdot (1 - \text{health})) \cdot (1 + 0.5 \cdot \rho)$ \\
Starvation & If energy $< 0.2$: $0.3 \cdot (1-R) \cdot (1 - \text{energy}/0.2)$ \\
Injury & 0.1 if injured \\
Old age & $0.01 \cdot (\text{age} - 50)$ if age $> 50$ \\
Crowding & $0.05 \cdot \max(0, \rho - 0.8) + \eta \cdot M_{\text{sink}}$ \\
\bottomrule
\end{tabular}
\end{table}

Here $\eta$ is the behavioral sink factor (Section~\ref{sec:sink}) and $M_{\text{sink}}$ is a binary mortality modifier active during sink engagement.

\subsection{Behavioral Sink and Recovery}
\label{sec:sink}

The behavioral sink is a density-driven phase transition that suppresses reproductive output even after density falls below carrying capacity.

\paragraph{Activation.}
When density exceeds the behavioral sink activation threshold (default 0.75), the sink engages. A sink factor is computed as a continuous measure of sink severity:

\begin{equation}
\eta = \min\left(1,\; \frac{\rho - 0.75}{1 - 0.75}\right)
\label{eq:sink_factor}
\end{equation}

\paragraph{Effects while active.}
Fertility is multiplicatively suppressed: $F \times (1 - \eta \cdot 0.85)$. Mating drive is suppressed: $M_d \times (1 - \eta \cdot 0.4)$. Withdrawal behavior and offspring neglect probabilities are increased.

\paragraph{Deactivation (hysteresis).}
The sink disengages only when density falls below the behavioral sink deactivation threshold (default 0.50), creating a hysteresis band (on at 0.75, off at 0.50). A minimum duration of 30 steps must elapse before deactivation is permitted. An optional auto-recovery timer (default 100 steps) forces disengagement after a fixed number of steps.

\paragraph{Post-sink recovery.}
After sink disengagement, a recovery phase lasting 100 steps begins. During recovery: fertility and mating drive are multiplicatively boosted by 1.3; mortality is reduced to 40\% of normal. If population drops below 30 during recovery, refugees (up to 10 per step) are injected to supplement the depleted population.

\paragraph{Sink Factor in the HSAP Composite Indicator.}
The composite indicator incorporates sink state: when the sink is active, the indicator is dampened, reflecting the disruption of normal social-endocrine regulation.

\subsection{Metrics}
\label{sec:metrics}

Per time step, HSAP records: population size, density, sex ratio, resource abundance, predator pressure, disease pressure, territory availability, age structure (juvenile/adult/senior counts), mean hormones (testosterone, estrogen, cortisol, male\_T, female\_T), behavioral outputs (male/female aggression, female defense), reproductive metrics (fertility, births, deaths, matings, pregnancies, infanticide, neglect, refugees), sink state (active, post-recovery, sink\_factor), external threat index, and the HSAP composite indicator.

The HSAP composite indicator is an analytical summary metric, not a directly measurable biological quantity. It is a weighted composite of five signals: low external threat (25\%), low male aggression (20\%), high female aggression (20\%), low fertility (20\%), and population stability (15\%), with a viability dampener for populations below 50. The weights were chosen to reflect the model's hypothesized causal chain; they are not fit to data.

\paragraph{Summary metrics} (computed over the full run): final population, peak population, minimum population, crash ratio, time to stability, mean hormones (testosterone, estrogen, cortisol), mean aggression (male, female), mean fertility, extinction status, and run number.

\subsection{Null Models}
\label{sec:null_models}

We compared HSAP against 11 null models that lack endocrine feedbacks or behavioral mechanisms. Each null model was fit to HSAP's population trajectory and evaluated by mean squared error (MSE). Full equations are provided in the supplementary materials.

\begin{table}[h]
\centering
\caption{Null model suite}
\label{tab:null_models}
\begin{tabular}{lll}
\toprule
Model & Mechanism & Key equation \\
\midrule
N0\_logistic & Logistic growth & $dN/dt = rN(1-N/K)$ \\
N1\_predator\_prey & Lotka-Volterra & $dN/dt = \alpha N - \beta NP$ \\
N2\_density\_fertility & Density-dependent fertility & $F = F_0 \cdot (1 - N/K)$ \\
N3\_disease\_pressure & Disease mortality & $M_d = D \cdot (1 + 0.5 \cdot N/K)$ \\
N4\_dominance\_hierarchy & Rank-based allocation & No endocrine feedback \\
N5\_random\_hormone & Random hormones & Hormones $\sim U(0,1)$ \\
N6\_resource\_only & Resource-limited growth & Growth limited by $R$ only \\
N7\_density\_recovery & Density + recovery & Density dependence + Allee term \\
N8\_random\_behavior & Random actions & Uniform action selection \\
N9\_social\_behavior & Social only & No endocrine coupling \\
N10\_endocrine\_no\_behavior & Endocrine only & Hormones do not affect behavior \\
\bottomrule
\end{tabular}
\end{table}

We also tested six HSAP ablation variants, each removing one component: male testosterone downshift, female aggression channel, cortisol, endocrine responsiveness, reproductive restraint, and sink recovery.

\subsection{Software and Reproducibility}
\label{sec:software}

The model is implemented in Python 3.10+ using NumPy, pandas, Matplotlib, and tqdm. Sensitivity analysis uses SALib. Genetic algorithm search uses DEAP. The model code and tests are available in the source repository.

All scenario definitions are frozen in the source repository. Simulation outputs are frozen in the results directory. The pipeline is reproducible via:

\begin{verbatim}
python scripts/run_resumable_pipeline.py validate
python scripts/run_resumable_pipeline.py summarize
python scripts/make_paper_figures.py --dpi 300
\end{verbatim}

\subsection{Parameterization and Model Calibration}
\label{sec:parameterization}

HSAP parameters were not estimated through optimization against empirical population trajectories. Instead, parameters define a computational experiment space used to examine whether coupled ecological and behavioral feedback mechanisms can generate emergent population dynamics. Biological literature informed the direction and plausibility of several mechanisms (e.g., testosterone suppresses aggression when threat is low; cortisol suppresses fertility), while numerical values function as model assumptions and scenario controls. The model contains 71 free parameters across six parameter classes (endocrine, behavioral, reproduction, mortality, environment, and GA). A full parameter table with provenance is provided in the supplementary materials.

This parameterization strategy is appropriate for the model's purpose: HSAP is a computational exploration, not a predictive statistical model. The question is not whether specific parameter values match empirical data, but whether the class of feedback structures described by HSAP can produce qualitatively distinct population dynamics under defined assumptions.

\subsection{Spatial Representation}
\label{sec:spatial}

The model uses a well-mixed population approximation. Agents are not assigned explicit spatial coordinates; interactions represent aggregate social dynamics rather than localized territorial processes. Territory availability is computed as a scalar function of density ($\tau = \max(0, 1 - \rho \cdot S)$), not as a spatially explicit resource landscape. This abstraction is appropriate for the model's scope: HSAP tests endocrine-behavioral feedback mechanisms, not spatial ecology.

\subsection{Interaction Model}
\label{sec:interaction}

Social interactions are represented as pairwise anonymous encounters rather than persistent network relationships. At each step, agents are processed in random order. Mating pairs are formed by selecting a male with probability proportional to mating drive and pairing with a randomly selected fertile female. Dominance contests occur when an agent selects the dominance event action; the opponent is not explicitly modeled (outcomes depend on the acting agent's rank and testosterone). There is no kin recognition, no reciprocal altruism, and no reputation system. Cooperation accumulates a bounded bonus ($\leq 0.3$) but does not create social bonds.

\subsection{Resource Scarcity}
\label{sec:resource_scarcity}

Resource scarcity is defined as:

\begin{equation}
\text{resource\_scarcity} = \max(0,\; 1 - R_t)
\label{eq:resource_scarcity}
\end{equation}

where $R_t$ is the current resource abundance. This scalar appears in the cortisol update, female testosterone update, aggression computation, offspring survival probability, and mortality equations. It represents the gap between current resource availability and maximum resource capacity, not a direct measure of starvation risk (which is modeled separately through the energy-based starvation mortality component).
