\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}
% \usepackage{lineno}
\usepackage{geometry}
\geometry{margin=1in}

\title{Hormonal-Social Adaptation Population Model:\\
A Computational Framework for Exploring Behavioral Stabilization Under Ecological Release}

\author{Jake Enholm\\
\textit{Data Scientist}\\
\small\url{https://github.com/jenholm/bio_models/tree/main/hsap}}

\date{2026}

\begin{document}

\maketitle

\begin{abstract}
Ecological conditions that reduce mortality pressure can alter the selective environment experienced by populations, yet most population regulation models focus on density-dependent resource competition and predation without incorporating endocrine-mediated behavioral feedbacks. We present HSAP (Hormonal-Social Adaptation Population model), an agent-based model testing whether coupled behavioral and endocrine mechanisms can influence population stability under reduced external threat. In HSAP, individuals regulate testosterone, cortisol, and estrogen in response to perceived predation risk and population density. These hormonal states drive behavioral outputs, including aggression, fertility, and mating drive, that feed back into population dynamics through a density-dependent behavioral sink mechanism with hysteresis. We evaluated 61 scenarios (4 core conditions, 3 sink variants, 54-factorial parameter sweep) across 1,970 validated simulation runs. Under reduced predation and disease pressure, populations stabilized at higher densities with reduced male testosterone and aggression (Cohen $d=3.29$). Under high space constraint with abundant resources, populations declined to 31\% of the resource-matched uncrowded condition, showing crowding pathology through fertility suppression. Under behavioral sink conditions, populations either recovered or produced partial collapse with 14\% extinction risk. A null model suite of 11 alternatives could not distinguish HSAP from logistic growth, predator-prey, or density-fertility models on population trajectories alone; HSAP became identifiable only when behavioral and endocrine observables were included. These results are consistent with the simulation assumptions: under the model's assumptions, endocrine-linked behavioral feedbacks can produce qualitatively distinct population dynamics under identical resource conditions, but population data alone are insufficient to identify the mechanistic pathways involved. The model does not establish that these mechanisms operate in natural populations; empirical validation requires paired endocrine and behavioral field measurements.
\end{abstract}

% \linenumbers

\input{sections/introduction}
\input{sections/methods}
\input{sections/results}
\input{sections/discussion}
\input{sections/conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
